---
title: "Combining R and Python with Reticulate"
author: "Ethan Liang"
date: "2026-09-28"
categories: [R, Python, Reticulate, Polyglot]
format:
  html:
    code-fold: false
    toc: true
---

## Introduction & Data Source

Data science teams frequently use both R and Python for their unique strengths—such as `ggplot2` for high-quality visualizations in R and `scikit-learn` for machine learning in Python. 

In this post, we demonstrate how to bridge both execution engines seamlessly within a single Quarto document using the `reticulate` package.

The dataset used is the **palmerpenguins** dataset, collected by Dr. Kristen Gorman and the Palmer Station LTER. You can read more on the [official palmerpenguins documentation page](https://allisonhorst.github.io/palmerpenguins/).

---

## 1. Data Preparation in R

We start in an **R chunk** by loading `palmerpenguins`, cleaning missing records, and computing species-level baseline aggregates.

```{r}
#| label: r-data-prep
#| message: false
#| warning: false
library(palmerpenguins)
library(dplyr)
library(tidyr)
library(reticulate)

# Clean dataset removing NAs across key physical measurements
penguins_clean <- penguins %>%
  drop_na(species, bill_length_mm, flipper_length_mm, body_mass_g) %>%
  mutate(species = as.character(species))

# Calculate summary metrics by species in R
species_summary <- penguins_clean %>%
  group_by(species) %>%
  summarise(
    n = n(),
    mean_flipper_mm = mean(flipper_length_mm),
    mean_mass_g = mean(body_mass_g),
    mean_bill_len_mm = mean(bill_length_mm),
    .groups = "drop"
  )

species_summary
```

## 2. Model Fitting in Python (Reading R Objects)

Python accesses that R summary object directly using r.species_summary, fits a linear regression model on raw penguin data passed from R, and computes predicted values.

```{python}
#| label: python-modeling
import pandas as pd
import numpy as np
from sklearn.linear_model import LinearRegression

# Access R summary object in Python
print("Accessing R species_summary inside Python:")
print(r.species_summary)

# Convert R dataframe to Python DataFrame
df_py = r.penguins_clean

# Define predictor features (X) and target variable (y)
X = df_py[['flipper_length_mm', 'bill_length_mm']]
y = df_py['body_mass_g']

# Fit Linear Regression model in Python
model = LinearRegression()
model.fit(X, y)

# Generate predictions and store summary metrics
df_py['predicted_mass_g'] = model.predict(X)
df_py['residual_g'] = df_py['body_mass_g'] - df_py['predicted_mass_g']

# Extract model coefficients for evaluation
r2_score = model.score(X, y)
coef_flipper = model.coef_[0]
coef_bill = model.coef_[1]
intercept = model.intercept_

print(f"\nModel Fit Complete (R^2 = {r2_score:.4f})")
print(f"Formula: Body Mass = {intercept:.2f} + {coef_flipper:.2f}*(Flipper Length) + {coef_bill:.2f}*(Bill Length)")
```

## 3. Visualizing Python Predictions Back in R

R retrieves the Python model results via py$predictions and renders the final comparative plot using ggplot2.

```{r}
#| label: fig-r-py-visualization
#| fig-cap: "Actual vs. Predicted Penguin Body Mass (Computed in Python, Rendered in R)."
#| fig-width: 8
#| fig-height: 5.5
library(ggplot2)

# Access Python object in R using py$...
penguin_preds <- py$df_py

ggplot(penguin_preds, aes(x = body_mass_g, y = predicted_mass_g, color = species)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray40", size = 0.8) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Actual vs. Predicted Body Mass Across Penguin Species",
    subtitle = "Model fit in Python (scikit-learn) and plotted in R (ggplot2)",
    x = "Actual Body Mass (g)",
    y = "Predicted Body Mass (g)",
    color = "Species"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```
